---
title: "Ball by ball Analysis"
output: pdf_document
date: "2026-04-02"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse,
               tidymodels,
               skimr,
               plotly,
               yardstick, 
               vip
               ,cricketdata)


df_ball_by_ball_data <- read_csv("./data/cleaned_ball_by_ball_data.csv")

#dropping the washed out or draw matches 
df_ball_by_ball_data <- df_ball_by_ball_data %>% 
  filter(!is.na(winner))


## final data used for creating models in Research Project part 1
df_part_1 <- read_csv("./data/final_df_part_1.csv")

```


#Ball by ball data 

## intital inspection   
```{r  initial_data_inspection_1}

skim_without_charts(df_ball_by_ball_data)

colnames(df_ball_by_ball_data)


```
```{r initial_data_inspection_2}
head(df_ball_by_ball_data)
```
## poweplay 
```{r}
# poweplay 
unique(df_ball_by_ball_data$over)
unique(df_ball_by_ball_data$powerplay_type)
```

```{r}
# sum(!df_ball_by_ball_data$runs_total>= df_ball_by_ball_data$runs_batter)
```

## Powerplay /Surge analysis 
```{r data_cleaning}
# Power surge rule first introduced date 
surge_intro_date <- as.Date("2020-12-10")

# Calculate match targets from first innings
match_targets <- df_ball_by_ball_data %>%
  group_by(match_id) %>%
  filter(innings == 1) %>%
  summarise(
    first_innings_total = sum(runs_total),
    target_to_win = first_innings_total + 1,
    .groups = "drop"
  )

# Aggregate ball-by-ball data to over-level 
df_over_analysis <- df_ball_by_ball_data %>%
  mutate(date = as.Date(date)) %>%
  group_by(match_id, innings, over) %>%
  
  summarise(
    # get the static labels using first()
    date   = first(date),
    team1   = first(team1),
    team2   = first(team2),
    winner = first(winner),  # overall match winner
    bowler = first(bowler),
    
    #calculations needed
    over_total_runs  = sum(runs_total),
    over_batter_runs = sum(runs_batter),
    over_extras      = sum(runs_total - runs_batter),
    over_wickets     = sum(wicket),
    boundaries       = sum(runs_batter %in% c(4, 6)),
    dot_balls        = sum(runs_total == 0),
    pp_type          = first(powerplay_type),
    venue  = first(venue),
    .groups = "drop"
  ) %>%
  mutate(
    era = ifelse(date < surge_intro_date, "Pre_Surge", "Surge_Era"),
    pp_type = replace_na(pp_type, "none")
  )

# # Define match phases using era-specific rules (6-over vs 4+2 system)
# df_over_analysis <- df_over_analysis %>%
#   mutate(match_phase = case_when(
#     era == "Pre_Surge" & over <= 6 ~ "Mandatory_PP",
#     era == "Surge_Era" & over <= 4 ~ "Mandatory_PP",
#     pp_type == "batting" ~ "Power_Surge",
#     over >= 16 ~ "Death",
#     TRUE ~ "Middle"
#   ))

df_over_analysis <- df_over_analysis %>%
  mutate(
    # Normal innings phase, regardless of Power Surge
    match_phase = case_when(
      era == "Pre_Surge" & over <= 6 ~ "Mandatory_PP",
      era == "Surge_Era" & over <= 4 ~ "Power_Surge_1",
      over >= 16 ~ "Death",
      TRUE ~ "Middle"
    ),
    
    # Tactical phase, where Power Surge is treated separately
    tactical_phase = case_when(
      era == "Pre_Surge" & over <= 6 ~ "Mandatory_PP",
      era == "Surge_Era" & over <= 4 ~ "Power_Surge_1",
      era == "Surge_Era" & pp_type == "batting" ~ "Power_Surge_2",
      over >= 16 ~ "Death",
      TRUE ~ "Middle"
    ),
    
    # When the Power Surge was taken
    time_when_surge_taken = case_when(
      tactical_phase == "Power_Surge_2" & match_phase == "Middle" ~ "Surge taken in Middle overs",
      tactical_phase == "Power_Surge_2" & match_phase == "Death" ~ "Surge taken in Death overs",
      TRUE ~ "No Power Surge"
    )
  )

# cumulative match state and current run rate
df_over_analysis <- df_over_analysis %>%
  arrange(match_id, innings, over) %>%
  group_by(match_id, innings) %>%
  mutate(
    current_score = cumsum(over_total_runs),
    wickets_lost = cumsum(over_wickets),
    current_run_rate = current_score / over
  ) %>%
  ungroup()

# Join targets and calculate Required Run Rate (RRR) for second innings
df_over_analysis <- df_over_analysis %>%
  left_join(match_targets, by = "match_id") %>%
  mutate(
    runs_remaining = ifelse(innings == 2, target_to_win - current_score, NA),
    overs_remaining = 20 - over,
    req_run_rate = case_when(
      innings == 1 ~ NA_real_,
      runs_remaining <= 0 ~ 0,
      overs_remaining == 0 ~ NA_real_,
      TRUE ~ runs_remaining / overs_remaining
    )
  )
```


##Creating bowling style metadata from existing ball by ball
```{r different_bowlers}
bowlers <- as_tibble(unique(df_ball_by_ball_data$bowler))
bowler_names <- unique(df_ball_by_ball_data$bowler)
# bowler_info <- map_df(bowler_names, find_player_id)
# write_csv(bowler_info, "./data/bowler_info.csv")

bowler_info <- read_csv("./data/bowler_info.csv")

colnames(bowler_info)

```

## Creating player metadata
####obtained from cricksheet https://cricsheet.org/register/
```{r creating_player_metadata}
# unique_ids <- unique(bowler_info$ID)
# detailed_meta <- fetch_player_meta(unique_ids)
# every call is expensive so just write it to a csv file and use it 
# player_meta_data <- write_csv(player_meta_data, "./data/player_meta_data.csv")

player_meta_data <- read_csv("./data/people_registry.csv")
player_meta_data <- player_meta_data %>% 
  rename(key_cricinfo = cricinfo_id)

#obtained from cricksheet https://cricsheet.org/register/
people <-  read_csv("./data/people.csv") 

people <- people %>% 
  select(unique_name, key_cricinfo)

my_bowlers <- people %>%
  filter(unique_name %in% bowler_names)
# To check if any of the bowlers are missing 
  # missing_bowlers <- tibble(unique_name = bowler_names) %>%
  #   anti_join(my_bowlers, by = "unique_name")
  # 
  # # View the missing players
  # print(missing_bowlers)

```
## Merging datsets (player Metadata & Bowling styles)
```{r merging_datasets}
colnames(player_meta_data)
colnames(my_bowlers)

bowler_type_map <- my_bowlers %>%
  # Clean join using the matching column name
  left_join(player_meta_data, by = "key_cricinfo") %>%
  select(key_cricinfo, unique_name, bowling_style) %>%
  mutate(bowling_category = case_when(
    str_detect(tolower(bowling_style), "fast|medium|seam|swing") ~ "Pace",
    str_detect(tolower(bowling_style), "spin|break|orthodox|leg|wrist") ~ "Spin",
    TRUE ~ "Unknown"
  ))
```


```{r adding_bowling_style_and_category}
# Combining Datasets Part 1 
# combining_match_and_ball_datasets

df_part_2 <- df_over_analysis %>%
  left_join(
    # Select only the necessary columns from the map to keep the dataset clean
    bowler_type_map %>% select(unique_name, bowling_style, bowling_category), 
    by = c("bowler" = "unique_name")
  )
skim_without_charts(df_over_analysis)
```

### creating over level dataset using bally by ball 
```{r creating_master_dataset}
# Combining Datasets Part 2 
# creating master dataset by merging datset from ResearchProject A to be used In part B 

# datasset(df_part_1 was used in research project Part A)
df_master <- df_part_2 %>%
  left_join(
    df_part_1 %>%
      select(
        match_id,
        season,
        toss_winner,
        toss_decision,
        home_team,
        home_wins,
        toss_win_game_win,
        toss_winner_home
      ),
    by = "match_id"
  )

df_before_surge <- df_master %>%
  mutate(date = as.Date(date)) %>%
  filter(date < surge_intro_date)

df_after_surge <- df_master %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= surge_intro_date)

colnames(df_master)
```

```{r}
skim_without_charts(df_master)
```

#EDA final
```{r}
over_profile <- df_master %>%
  group_by(innings, over) %>%
  summarise(
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    avg_wickets = mean(over_wickets, na.rm = TRUE),
    avg_boundaries = mean(boundaries, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(over_profile, aes(x = over, y = avg_runs, color = factor(innings))) +
  geom_line(linewidth = 1.1) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Runs per Over by Innings",
    x = "Over",
    y = "Average Runs",
    color = "Innings"
  )


# Before vs after Power Surge by over
surge_over_profile <- df_master %>%
  group_by(era, over) %>%
  summarise(
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    .groups = "drop"
  )


ggplot(surge_over_profile, aes(x = over, y = avg_runs, color = era)) +
  geom_line(linewidth = 1.1) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Runs per Over Before and After Power Surge Introduction",
    x = "Over",
    y = "Average Runs",
    color = "Era"
  )


# Boundary rate and dot-ball rate by match phase
phase_style <- df_master %>%
  group_by(match_phase) %>%
  summarise(
    avg_boundaries = mean(boundaries, na.rm = TRUE),
    avg_dot_balls = mean(dot_balls, na.rm = TRUE),
    .groups = "drop"
  )

phase_style


# Boundary rate and dot-ball rate by tactical phase
tactical_phase_style <- df_master %>%
  group_by(tactical_phase) %>%
  summarise(
    avg_boundaries = mean(boundaries, na.rm = TRUE),
    avg_dot_balls = mean(dot_balls, na.rm = TRUE),
    .groups = "drop"
  )

tactical_phase_style


# Death over performance including Power Surge
death_over_performance <- df_master %>%
  filter(match_phase == "Death") %>%
  group_by(era) %>%
  summarise(
    death_overs = n(),
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    avg_wickets = mean(over_wickets, na.rm = TRUE),
    avg_boundaries = mean(boundaries, na.rm = TRUE),
    avg_dot_balls = mean(dot_balls, na.rm = TRUE),
    .groups = "drop"
  )

death_over_performance


# Death over performance excluding Power Surge
normal_death_performance <- df_master %>%
  filter(tactical_phase == "Death") %>%
  group_by(era) %>%
  summarise(
    death_overs = n(),
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    avg_wickets = mean(over_wickets, na.rm = TRUE),
    avg_boundaries = mean(boundaries, na.rm = TRUE),
    avg_dot_balls = mean(dot_balls, na.rm = TRUE),
    .groups = "drop"
  )

normal_death_performance
```

<!--
Logic needs to be changed for EDA since phase level logic was flawed 
#EDA V1 wickets 
```{r }
unique(df_master$match_phase)
unique(df_master$pp_type)
# Runs by match phase
df_master %>%
  group_by(match_phase) %>%
  summarise(
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    avg_wickets = mean(over_wickets, na.rm = TRUE),
    avg_boundaries = mean(boundaries, na.rm = TRUE)
  )

df_master %>%
  group_by(match_phase, bowling_category) %>%
  summarise(
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    avg_wickets = mean(over_wickets, na.rm = TRUE),
    .groups = "drop"
  )

# Since the Power Surge rule was introduced on 10 December 2020, the exploratory analysis compares matches played before and after this date. Before the rule change innings were divided into three phases: Mandatory Powerplay, Middle, and Death. After the introduction of the Power Surge, a fourth phase, Power Surge, was added. Therefore, the analysis examines whether scoring patterns, wicket loss, boundary frequency, and bowling effectiveness changed across phases between the two periods.

# Power Surge introduction date
surge_intro_date <- as.Date("2020-12-10")

df_eda <- df_master %>%
  mutate(
    date = as.Date(date),
    surge_period = case_when(
      date < surge_intro_date ~ "Before Power Surge",
      TRUE ~ "After Power Surge"
    )
  )

# Summary of scoring by phase and period
bowling_phase_period_summary <- df_eda %>%
  group_by(surge_period, match_phase, bowling_category) %>%
  summarise(
    overs = n(),
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    avg_wickets = mean(over_wickets, na.rm = TRUE),
    .groups = "drop"
  )

# Visualise bowling type effect on runs
ggplot(bowling_phase_period_summary,
       aes(x = match_phase, y = avg_runs, fill = bowling_category)) +
  geom_col(position = "dodge") +
  facet_wrap(~ surge_period) +
  theme_minimal() +
  labs(
    title = "Average Runs per Over by Bowling Type and Match Phase",
    x = "Match Phase",
    y = "Average Runs per Over",
    fill = "Bowling Type"
  )

# Visualise bowling type effect on wickets
ggplot(bowling_phase_period_summary,
       aes(x = match_phase, y = avg_wickets, fill = bowling_category)) +
  geom_col(position = "dodge") +
  facet_wrap(~ surge_period) +
  theme_minimal() +
  labs(
    title = "Average Wickets per Over by Bowling Type and Match Phase",
    x = "Match Phase",
    y = "Average Wickets per Over",
    fill = "Bowling Type"
  )
```


# Eda v2 Average runs Scored 
```{r }
over_profile <- df_master %>%
  group_by(innings, over) %>%
  summarise(
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    avg_wickets = mean(over_wickets, na.rm = TRUE),
    avg_boundaries = mean(boundaries, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(over_profile, aes(x = over, y = avg_runs, color = factor(innings))) +
  geom_line(linewidth = 1.1) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Runs per Over by Innings",
    x = "Over",
    y = "Average Runs",
    color = "Innings"
  )


# Before vs after Power Surge by over
surge_over_profile <- df_master %>%
  mutate(
    surge_period = if_else(
      as.Date(date) < as.Date("2020-12-10"),
      "Before Power Surge",
      "After Power Surge"
    )
  ) %>%
  group_by(surge_period, over) %>%
  summarise(
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    .groups = "drop"
  )


ggplot(surge_over_profile, aes(x = over, y = avg_runs, color = surge_period)) +
  geom_line(linewidth = 1.1) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Runs per Over Before and After Power Surge Introduction",
    x = "Over",
    y = "Average Runs",
    color = "Period"
  )


# Boundary rate and dot-ball rate by phase
phase_style <- df_master %>%
  group_by(match_phase) %>%
  summarise(
    avg_boundaries = mean(boundaries, na.rm = TRUE),
    avg_dot_balls = mean(dot_balls, na.rm = TRUE),
    .groups = "drop"
  )

phase_style
```
-->

```{r}
colnames(df_master)
```

<!-- #---------------------------------------------------------------------------------------------- -->

# Modeling 

## Section 1
Improvement of part A by using ball-by-ball information

#### Checkpoint logic 
```{r checkpoint_logic, include=FALSE}
# Create phase-aware checkpoint dataset
df_section_1 <- df_master %>%
  arrange(match_id, innings, over) %>%
  group_by(match_id, innings) %>%
  mutate(
    end_power_surge_1 = if_else(
      any(tactical_phase == "Power_Surge_1"),
      max(over[tactical_phase == "Power_Surge_1"], na.rm = TRUE),
      NA_real_
    ),

    end_mandatory_pp = if_else(
      any(tactical_phase == "Mandatory_PP"),
      max(over[tactical_phase == "Mandatory_PP"], na.rm = TRUE),
      NA_real_
    ),

    end_power_surge_2 = if_else(
      any(tactical_phase == "Power_Surge_2"),
      max(over[tactical_phase == "Power_Surge_2"], na.rm = TRUE),
      NA_real_
    ),

    checkpoint_type = case_when(
      era == "Pre_Surge" & over == end_mandatory_pp ~ "End_Mandatory_PP",
      era == "Surge_Era" & over == end_power_surge_1 ~ "End_Power_Surge_1",
      era == "Surge_Era" & over == end_power_surge_2 ~ "End_Power_Surge_2",
      over == 10 ~ "Mid_Innings",
      over == 16 ~ "Death_Start",
      TRUE ~ NA_character_
    )
  ) %>%
  ungroup() %>%
  filter(!is.na(checkpoint_type)) %>%
  mutate(
    toss_loser = case_when(
      toss_winner == team1 ~ team2,
      toss_winner == team2 ~ team1,
      TRUE ~ NA_character_
    ),

    batting_team = case_when(
      toss_decision == "bat" & innings == 1 ~ toss_winner,
      toss_decision == "bat" & innings == 2 ~ toss_loser,
      toss_decision == "field" & innings == 1 ~ toss_loser,
      toss_decision == "field" & innings == 2 ~ toss_winner,
      TRUE ~ NA_character_
    ),

    bowling_team = case_when(
      batting_team == team1 ~ team2,
      batting_team == team2 ~ team1,
      TRUE ~ NA_character_
    ),

    toss_winner_batting = factor(
      if_else(batting_team == toss_winner, "Yes", "No"),
      levels = c("No", "Yes")
    ),

    toss_win_game_win = factor(
      if_else(winner == toss_winner, "Yes", "No"),
      levels = c("No", "Yes")
    ),

    innings = factor(innings),
    checkpoint_type = factor(
      checkpoint_type,
      levels = c(
        "End_Mandatory_PP",
        "End_Power_Surge_1",
        "End_Power_Surge_2",
        "Mid_Innings",
        "Death_Start"
      )
    ),
    toss_decision = factor(toss_decision, levels = c("bat", "field")),
    toss_winner_home = factor(toss_winner_home, levels = c("away", "home", "neutral")),
    toss_winner = factor(toss_winner),
    toss_loser = factor(toss_loser),
    batting_team = factor(batting_team),
    bowling_team = factor(bowling_team),
    match_phase = factor(match_phase),
    tactical_phase = factor(tactical_phase),
    era = factor(era),
    venue = factor(venue),
    season = factor(season)
  ) %>%
  select(
    match_id,
    date,
    era,
    innings,
    over,
    checkpoint_type,
    toss_win_game_win,
    toss_decision,
    toss_winner_home,
    toss_winner,
    toss_loser,
    batting_team,
    bowling_team,
    toss_winner_batting,
    match_phase,
    tactical_phase,
    time_when_surge_taken,
    current_score,
    wickets_lost,
    current_run_rate
  ) %>%
  drop_na(toss_win_game_win)

# Check checkpoint distribution
df_section_1 %>%
  count(era, checkpoint_type) %>%
  print(n = Inf)

# Check batting logic
df_section_1 %>%
  count(toss_decision, innings, toss_winner_batting) %>%
  print(n = Inf)

# Check outcome balance
df_section_1 %>%
  count(toss_win_game_win)

# Check checkpoints by innings
df_section_1 %>%
  count(innings, checkpoint_type) %>%
  print(n = Inf)
```


### Model A, B, C 
```{r part_b_section1_partA, include=FALSE}
set.seed(42)

# Match-level train/test split
match_ids <- unique(df_section_1$match_id)

train_ids <- sample(
  match_ids,
  size = floor(0.8 * length(match_ids))
)

test_ids <- setdiff(match_ids, train_ids)

section1_train <- df_section_1 %>%
  filter(match_id %in% train_ids)

section1_test <- df_section_1 %>%
  filter(match_id %in% test_ids)

section1_metrics <- metric_set(
  accuracy,
  roc_auc,
  brier_class
)

 # Helper function
 fit_eval_logit <- function(model_formula, train_data, test_data, model_name) {

  rec <- recipe(model_formula, data = train_data) %>%
    step_nzv(all_predictors())

  mod <- logistic_reg() %>%
    set_engine("glm") %>%
    set_mode("classification")

  wf <- workflow() %>%
    add_model(mod) %>%
    add_recipe(rec)

  fit_obj <- fit(wf, data = train_data)

  preds <- predict(fit_obj, new_data = test_data, type = "prob") %>%
    bind_cols(predict(fit_obj, new_data = test_data, type = "class")) %>%
    bind_cols(test_data %>% select(toss_win_game_win))

  metrics_tbl <- section1_metrics(
    preds,
    truth = toss_win_game_win,
    estimate = .pred_class,
    .pred_Yes
  ) %>%
    mutate(model = model_name)

  list(
    name = model_name,
    fit = fit_obj,
    preds = preds,
    metrics = metrics_tbl,
    aic = AIC(extract_fit_engine(fit_obj))
  )
}

# Model A: Part A baseline
model_A_formula <- toss_win_game_win ~
  toss_decision +
  toss_winner_home

model_A <- fit_eval_logit(
  model_A_formula,
  section1_train,
  section1_test,
  "Model A: Part A baseline"
)

# Model B: Part A baseline with team fixed effects
model_B_formula <- toss_win_game_win ~
  toss_decision +
  toss_winner_home +
  toss_winner +
  toss_loser

model_B <- fit_eval_logit(
  model_B_formula,
  section1_train,
  section1_test,
  "Model B: Team fixed effects"
)

# Model C: Team fixed effects with phase-aware ball-by-ball state
model_C_formula <- toss_win_game_win ~
  toss_decision +
  toss_winner_home +
  toss_winner +
  toss_loser +
  innings +
  checkpoint_type +
  toss_winner_batting +
  current_score +
  wickets_lost +
  current_run_rate

model_C <- fit_eval_logit(
  model_C_formula,
  section1_train,
  section1_test,
  "Model C: Team FE + phase-aware state"
)

# Model comparison
section1_comparison <- bind_rows(
  model_A$metrics,
  model_B$metrics,
  model_C$metrics
) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(
    names_from = .metric,
    values_from = .estimate
  ) %>%
  left_join(
    tibble(
      model = c(
        model_A$name,
        model_B$name,
        model_C$name
      ),
      AIC = c(
        model_A$aic,
        model_B$aic,
        model_C$aic
      )
    ),
    by = "model"
  )

print(section1_comparison)

# ROC comparison
roc_A <- roc_curve(model_A$preds, truth = toss_win_game_win, .pred_Yes) %>%
  mutate(model = "A: Part A baseline")

roc_B <- roc_curve(model_B$preds, truth = toss_win_game_win, .pred_Yes) %>%
  mutate(model = "B: Team fixed effects")

roc_C <- roc_curve(model_C$preds, truth = toss_win_game_win, .pred_Yes) %>%
  mutate(model = "C: Team FE + state")

bind_rows(roc_A, roc_B, roc_C) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(linewidth = 1.1) +
  geom_abline(linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "ROC Comparison",
    subtitle = "Improvement of Part A using phase-aware ball-by-ball checkpoints",
    x = "1 - Specificity",
    y = "Sensitivity",
    color = "Model"
  )

# Model summaries
summary(extract_fit_engine(model_A$fit))
summary(extract_fit_engine(model_B$fit))
summary(extract_fit_engine(model_C$fit))

print(section1_comparison)
```


### Model D mixed effect 
#### Setup
```{r section1_mixed_effects_setup, message=FALSE, warning=FALSE, include=FALSE}
# Model D: Mixed-effects logistic regression using lme4

pacman::p_load(lme4, broom.mixed)

# Prepare train data for lme4
section1_train_mixed <- section1_train %>%
  mutate(
    toss_win_game_win_num = if_else(toss_win_game_win == "Yes", 1, 0),
    toss_decision = factor(toss_decision, levels = c("bat", "field")),
    toss_winner_home = factor(toss_winner_home, levels = c("away", "home", "neutral")),
    innings = factor(innings),
    checkpoint_type = factor(checkpoint_type),
    toss_winner_batting = factor(toss_winner_batting, levels = c("No", "Yes")),
    toss_winner = factor(toss_winner),
    toss_loser = factor(toss_loser)
  ) %>%
  drop_na(
    toss_win_game_win_num,
    toss_decision,
    toss_winner_home,
    innings,
    checkpoint_type,
    toss_winner_batting,
    current_score,
    wickets_lost,
    current_run_rate,
    toss_winner,
    toss_loser
  )

# Prepare test data for lme4
section1_test_mixed <- section1_test %>%
  mutate(
    toss_win_game_win_num = if_else(toss_win_game_win == "Yes", 1, 0),
    toss_decision = factor(toss_decision, levels = c("bat", "field")),
    toss_winner_home = factor(toss_winner_home, levels = c("away", "home", "neutral")),
    innings = factor(innings),
    checkpoint_type = factor(checkpoint_type),
    toss_winner_batting = factor(toss_winner_batting, levels = c("No", "Yes")),
    toss_winner = factor(toss_winner),
    toss_loser = factor(toss_loser)
  ) %>%
  drop_na(
    toss_win_game_win_num,
    toss_decision,
    toss_winner_home,
    innings,
    checkpoint_type,
    toss_winner_batting,
    current_score,
    wickets_lost,
    toss_winner,
    toss_loser
  )
```

#### Model
```{r section1_mixed_effects_model, message=FALSE, warning=FALSE}
# Fit mixed-effects logistic regression
model_D_mixed <- lme4::glmer(
  toss_win_game_win_num ~
    toss_decision +
    toss_winner_home +
    innings +
    checkpoint_type +
    toss_winner_batting +
    current_score +
    wickets_lost +
    (1 | toss_winner) +
    (1 | toss_loser),
  family = binomial(link = "logit"),
  data = section1_train_mixed
)
```


```{r section1_me_performance, message=FALSE, warning=FALSE}
# Fixed effect odds ratios
mixed_odds_ratios <- broom.mixed::tidy(
  model_D_mixed,
  effects = "fixed",
  exponentiate = TRUE,
  conf.int = TRUE
)

print("Fixed effect odds ratios")
mixed_odds_ratios

# Random effect variance components

mixed_random_effects <- broom.mixed::tidy(
  model_D_mixed,
  effects = "ran_pars"
)

print("Random effect variance components")
mixed_random_effects

# Predict on test data
mixed_pred_prob <- predict(
  model_D_mixed,
  newdata = section1_test_mixed,
  type = "response",
  allow.new.levels = TRUE
)

mixed_preds <- section1_test_mixed %>%
  mutate(
    .pred_Yes = mixed_pred_prob,
    .pred_No = 1 - .pred_Yes,
    .pred_class = factor(
      if_else(.pred_Yes >= 0.5, "Yes", "No"),
      levels = c("No", "Yes")
    ),
    toss_win_game_win = factor(
      if_else(toss_win_game_win_num == 1, "Yes", "No"),
      levels = c("No", "Yes")
    )
  )

# Mixed model metrics
mixed_metrics <- section1_metrics(
  mixed_preds,
  truth = toss_win_game_win,
  estimate = .pred_class,
  .pred_Yes
) %>%
  mutate(model = "Model D: Mixed effects")

print("Predict on test data")
mixed_metrics

# Add mixed model to previous models A,b,c in a  comparison table
section1_comparison_mixed <- bind_rows(
  section1_comparison %>%
    select(model, accuracy, roc_auc, brier_class) %>%
    pivot_longer(
      cols = c(accuracy, roc_auc, brier_class),
      names_to = ".metric",
      values_to = ".estimate"
    ),
  mixed_metrics
) %>%
  pivot_wider(
    names_from = .metric,
    values_from = .estimate
  ) %>%
  left_join(
    tibble(
      model = c(
        section1_comparison$model,
        "Model D: Mixed effects"
      ),
      AIC = c(
        section1_comparison$AIC,
        AIC(model_D_mixed)
      )
    ),
    by = "model"
  )

section1_comparison_mixed

roc_mixed <- roc_curve(mixed_preds, truth = toss_win_game_win, .pred_Yes) %>%
  mutate(model = "D: Mixed effects")

# ROC comparison including mixed model
bind_rows(roc_A, roc_B, roc_C, roc_mixed) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(linewidth = 1.1) +
  geom_abline(linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "ROC Comparison with Mixed Effects",
    subtitle = "Team strength modelled using lme4 random intercepts",
    x = "1 - Specificity",
    y = "Sensitivity",
    color = "Model"
  )
```

```{r}
summary(model_D_mixed)
```




## Section 2
Bowling tactics by phase of the match

```{r section2_bowling_tactics_setup, message=FALSE, warning=FALSE}
# Section 2 dataset
df_section_2 <- df_master %>%
  mutate(
    toss_loser = case_when(
      toss_winner == team1 ~ team2,
      toss_winner == team2 ~ team1,
      TRUE ~ NA_character_
    ),

    batting_team = case_when(
      toss_decision == "bat" & innings == 1 ~ toss_winner,
      toss_decision == "bat" & innings == 2 ~ toss_loser,
      toss_decision == "field" & innings == 1 ~ toss_loser,
      toss_decision == "field" & innings == 2 ~ toss_winner,
      TRUE ~ NA_character_
    ),

    bowling_team = case_when(
      batting_team == team1 ~ team2,
      batting_team == team2 ~ team1,
      TRUE ~ NA_character_
    ),

    innings = factor(innings),
    tactical_phase = factor(
      tactical_phase,
      levels = c(
        "Mandatory_PP",
        "Power_Surge_1",
        "Power_Surge_2",
        "Middle",
        "Death"
      )
    ),
    bowling_category = factor(bowling_category, levels = c("Pace", "Spin")),
    bowler = factor(bowler),
    match_id = factor(match_id),
    batting_team = factor(batting_team),
    bowling_team = factor(bowling_team)
  ) %>%
  filter(
    bowling_category %in% c("Pace", "Spin"),
    !is.na(tactical_phase),
    !is.na(over_total_runs),
    !is.na(bowler),
    !is.na(match_id)
  )

# Descriptive summary
section2_summary <- df_section_2 %>%
  group_by(tactical_phase, bowling_category) %>%
  summarise(
    overs = n(),
    avg_runs = mean(over_total_runs, na.rm = TRUE),
    avg_wickets = mean(over_wickets, na.rm = TRUE),
    avg_boundaries = mean(boundaries, na.rm = TRUE),
    avg_dot_balls = mean(dot_balls, na.rm = TRUE),
    .groups = "drop"
  )

# Visualise average runs by bowling type and tactical phase
ggplot(
  section2_summary,
  aes(x = tactical_phase, y = avg_runs, fill = bowling_category)
) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Average Runs Conceded by Bowling Type and Tactical Phase",
    x = "Tactical Phase",
    y = "Average Runs per Over",
    fill = "Bowling Type"
  )
```


```{r section2_bowling_tactics_model, message=FALSE, warning=FALSE}
# Model 2: Linear Mixed-effects model for runs conceded
model_2_ME <- lmer(
  over_total_runs ~ bowling_category * tactical_phase + innings +
    (1 | match_id) +
    (1 | bowler),
  data = df_section_2
)

# Model summary
summary(model_2_ME)

# Fixed effect estimates
section2_fixed_effects <- broom.mixed::tidy(
  model_2_ME,
  effects = "fixed",
  conf.int = TRUE
)

# Random effect variance components
section2_random_effects <- broom.mixed::tidy(
  model_2_ME,
  effects = "ran_pars"
)

# Check if singularity is reached in this model 
isSingular(model_2_ME)

# Final outputs for review
print(section2_summary)
print(section2_fixed_effects)
print(section2_random_effects)
```

#### spin_vs_pace_difference in differernt phases 

```{r}
# Estimated phase-specific spin vs pace difference
spin_phase_effects <- tibble(
  phase = c(
    "Mandatory_PP",
    "Power_Surge_1",
    "Power_Surge_2",
    "Middle",
    "Death"
  ),
  spin_vs_pace_difference = c(
    fixef(model_2_ME)["bowling_categorySpin"],
    fixef(model_2_ME)["bowling_categorySpin"] +
      fixef(model_2_ME)["bowling_categorySpin:tactical_phasePower_Surge_1"],
    fixef(model_2_ME)["bowling_categorySpin"] +
      fixef(model_2_ME)["bowling_categorySpin:tactical_phasePower_Surge_2"],
    fixef(model_2_ME)["bowling_categorySpin"] +
      fixef(model_2_ME)["bowling_categorySpin:tactical_phaseMiddle"],
    fixef(model_2_ME)["bowling_categorySpin"] +
      fixef(model_2_ME)["bowling_categorySpin:tactical_phaseDeath"]
  )
)

spin_phase_effects
```

#### Based on the above results predicting match outcomes

```{r section2_match_outcome_model, message=FALSE, warning=FALSE}
set.seed(42)

spin_preferred_phases <- c(
  "Mandatory_PP",
  "Power_Surge_2",
  "Middle",
  "Death"
)

df_section_2_outcome <- df_section_2 %>%
  mutate(
    bowling_team_win = factor(
      if_else(bowling_team == winner, "Yes", "No"),
      levels = c("No", "Yes")
    ),
    is_spin_preferred_phase = tactical_phase %in% spin_preferred_phases,
    spin_used_in_preferred_phase = if_else(
      bowling_category == "Spin" & is_spin_preferred_phase,
      1,
      0
    )
  ) %>%
  group_by(match_id, bowling_team) %>%
  summarise(
    bowling_team_win = first(bowling_team_win),
    spin_preferred_phase_overs = sum(spin_used_in_preferred_phase, na.rm = TRUE),
    preferred_phase_runs_conceded = sum(
      if_else(is_spin_preferred_phase, over_total_runs, 0),
      na.rm = TRUE
    ),
    preferred_phase_wickets = sum(
      if_else(is_spin_preferred_phase, over_wickets, 0),
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  drop_na(bowling_team_win)

df_section_2_outcome %>%
  count(bowling_team_win)

outcome_model <- glm(
  bowling_team_win ~
    spin_preferred_phase_overs +
    preferred_phase_runs_conceded +
    preferred_phase_wickets,
  data = df_section_2_outcome,
  family = binomial(link = "logit")
)
```

### performance 
```{r section2_match_outcome_performance, message=FALSE, warning=FALSE}
summary(outcome_model)

broom::tidy(
  outcome_model,
  exponentiate = TRUE,
  conf.int = TRUE
)

outcome_split <- initial_split(
  df_section_2_outcome,
  prop = 0.8,
  strata = bowling_team_win
)

outcome_train <- training(outcome_split)
outcome_test  <- testing(outcome_split)

outcome_fit <- glm(
  bowling_team_win ~
    spin_preferred_phase_overs +
    preferred_phase_runs_conceded +
    preferred_phase_wickets,
  data = outcome_train,
  family = binomial(link = "logit")
)

outcome_preds <- outcome_test %>%
  mutate(
    .pred_Yes = predict(
      outcome_fit,
      newdata = outcome_test,
      type = "response"
    ),
    .pred_No = 1 - .pred_Yes,
    .pred_class = factor(
      if_else(.pred_Yes >= 0.5, "Yes", "No"),
      levels = c("No", "Yes")
    )
  )

outcome_metrics <- metric_set(
  accuracy,
  roc_auc,
  brier_class
)

performance <- outcome_metrics(
  outcome_preds,
  truth = bowling_team_win,
  estimate = .pred_class,
  .pred_Yes,
  event_level = "second"
)

performance

roc_curve(
  outcome_preds,
  truth = bowling_team_win,
  .pred_Yes,
  event_level = "second"
) %>%
  autoplot() +
  ggtitle("Section 2 Inference Outcome Model — ROC")
```

## Section 3
Momentum decision and match outcomes 

### Setup
```{r section3_momentum_setup, message=FALSE, warning=FALSE}
pacman::p_load(slider)

# Create Section 3 dataset
df_section_3 <- df_master %>%
  arrange(match_id, innings, over) %>%
  group_by(match_id, innings) %>%
  mutate(
    runs_last_3 = slider::slide_dbl(
      over_total_runs,
      sum,
      .before = 2,
      .complete = FALSE
    ),
    wickets_last_3 = slider::slide_dbl(
      over_wickets,
      sum,
      .before = 2,
      .complete = FALSE
    ),
    last_wicket_over = cummax(if_else(over_wickets > 0, over, 0)),
    overs_since_last_wicket = over - last_wicket_over
  ) %>%
  ungroup() %>%
  mutate(
    toss_loser = case_when(
      toss_winner == team1 ~ team2,
      toss_winner == team2 ~ team1,
      TRUE ~ NA_character_
    ),

    batting_team = case_when(
      toss_decision == "bat" & innings == 1 ~ toss_winner,
      toss_decision == "bat" & innings == 2 ~ toss_loser,
      toss_decision == "field" & innings == 1 ~ toss_loser,
      toss_decision == "field" & innings == 2 ~ toss_winner,
      TRUE ~ NA_character_
    ),

    bowling_team = case_when(
      batting_team == team1 ~ team2,
      batting_team == team2 ~ team1,
      TRUE ~ NA_character_
    ),

    batting_team_win = factor(
      if_else(batting_team == winner, "Yes", "No"),
      levels = c("No", "Yes")
    ),

    batting_team_win_num = if_else(batting_team_win == "Yes", 1, 0),

    innings = factor(innings),
    tactical_phase = factor(tactical_phase),
    batting_team = factor(batting_team),
    bowling_team = factor(bowling_team),
    match_id = factor(match_id)
  ) %>%
  filter(
    !is.na(batting_team_win_num),
    !is.na(current_score),
    !is.na(wickets_lost),
    !is.na(current_run_rate),
    !is.na(runs_last_3),
    !is.na(wickets_last_3),
    !is.na(overs_since_last_wicket),
    !is.na(batting_team),
    !is.na(bowling_team),
    !is.na(match_id),
    !is.na(tactical_phase)
  )

# Basic checks
df_section_3 %>%
  count(batting_team_win) 

df_section_3 %>%
  group_by(batting_team_win) %>%
  summarise(
    rows = n(),
    avg_current_score = mean(current_score, na.rm = TRUE),
    avg_wickets_lost = mean(wickets_lost, na.rm = TRUE),
    avg_current_run_rate = mean(current_run_rate, na.rm = TRUE),
    avg_runs_last_3 = mean(runs_last_3, na.rm = TRUE),
    avg_wickets_last_3 = mean(wickets_last_3, na.rm = TRUE),
    avg_overs_since_last_wicket = mean(overs_since_last_wicket, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

### Split
```{r section3_momentum_data_split, message=FALSE, warning=FALSE}
set.seed(42)

section3_match_ids <- unique(df_section_3$match_id)

section3_train_ids <- sample(
  section3_match_ids,
  size = floor(0.8 * length(section3_match_ids))
)

section3_test_ids <- setdiff(section3_match_ids, section3_train_ids)

section3_train_raw <- df_section_3 %>%
  filter(match_id %in% section3_train_ids)

section3_test_raw <- df_section_3 %>%
  filter(match_id %in% section3_test_ids)

section3_numeric_vars <- c(
  "current_score",
  "wickets_lost",
  "current_run_rate",
  "runs_last_3",
  "wickets_last_3",
  "overs_since_last_wicket"
)

section3_scaling_values <- section3_train_raw %>%
  summarise(
    across(
      all_of(section3_numeric_vars),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        sd = ~ sd(.x, na.rm = TRUE)
      )
    )
  )

scale_section3_data <- function(data, scaling_values) {
  data %>%
    mutate(
      current_score_z =
        (current_score - scaling_values$current_score_mean) /
        scaling_values$current_score_sd,
      
      wickets_lost_z =
        (wickets_lost - scaling_values$wickets_lost_mean) /
        scaling_values$wickets_lost_sd,
      
      current_run_rate_z =
        (current_run_rate - scaling_values$current_run_rate_mean) /
        scaling_values$current_run_rate_sd,
      
      runs_last_3_z =
        (runs_last_3 - scaling_values$runs_last_3_mean) /
        scaling_values$runs_last_3_sd,
      
      wickets_last_3_z =
        (wickets_last_3 - scaling_values$wickets_last_3_mean) /
        scaling_values$wickets_last_3_sd,
      
      overs_since_last_wicket_z =
        (overs_since_last_wicket - scaling_values$overs_since_last_wicket_mean) /
        scaling_values$overs_since_last_wicket_sd
    )
}

section3_train <- scale_section3_data(
  section3_train_raw,
  section3_scaling_values
)

section3_test <- scale_section3_data(
  section3_test_raw,
  section3_scaling_values
)
```

### Model 

```{r section3_momentum_model, message=FALSE, warning=FALSE}
# Model 3A: predicting outcome based on current state of the match
model_3A_current_state_only <- lme4::glmer(
  batting_team_win_num ~
    innings +
    tactical_phase +
    current_score_z +
    wickets_lost_z +
    current_run_rate_z +
    (1 | batting_team) +
    (1 | bowling_team) +
    (1 | match_id),
  data = section3_train,
  family = binomial(link = "logit"),
  
)

summary(model_3A_current_state_only)

# Model 3B: predicting outcome based on current state of the match + recent momentum 
model_3B_momentum <- lme4::glmer(
  batting_team_win_num ~
    innings +
    tactical_phase +
    current_score_z +
    wickets_lost_z +
    current_run_rate_z +
    runs_last_3_z +
    wickets_last_3_z +
    overs_since_last_wicket_z +
    (1 | batting_team) +
    (1 | bowling_team) +
    (1 | match_id),
  data = section3_train,
  family = binomial(link = "logit"),
)

summary(model_3B_momentum)

# Fixed effect odds ratios
section3_state_only_or <- broom.mixed::tidy(
  model_3A_current_state_only,
  effects = "fixed",
  exponentiate = TRUE,
  conf.int = TRUE
)

section3_momentum_or <- broom.mixed::tidy(
  model_3B_momentum,
  effects = "fixed",
  exponentiate = TRUE,
  conf.int = TRUE
)

section3_state_only_or
section3_momentum_or

# Random effect variance components
section3_state_random <- broom.mixed::tidy(
  model_3A_current_state_only,
  effects = "ran_pars"
)

section3_momentum_random <- broom.mixed::tidy(
  model_3B_momentum,
  effects = "ran_pars"
)

section3_state_random
section3_momentum_random

# Model comparison
section3_model_comparison <- tibble(
  model = c(
    "Model 3A: Current State only",
    "Model 3B: Current State + momentum"
  ),
  AIC = c(
    AIC(model_3A_current_state_only),
    AIC(model_3B_momentum)
  ),
  BIC = c(
    BIC(model_3A_current_state_only),
    BIC(model_3B_momentum)
  ),
  logLik = c(
    as.numeric(logLik(model_3A_current_state_only)),
    as.numeric(logLik(model_3B_momentum))
  ),
  singular = c(
    isSingular(model_3A_current_state_only),
    isSingular(model_3B_momentum)
  )
)

section3_model_comparison

# Final outputs
print(section3_model_comparison)
# print(section3_state_only_or, n = Inf)
# print(section3_momentum_or, n = Inf)
print(section3_state_random)
print(section3_momentum_random)
```

### performance

```{r section3_performance, message=FALSE, warning=FALSE}
section3_metrics <- metric_set(
  accuracy,
  roc_auc,
  brier_class
)

make_section3_preds <- function(model, data) {
  data %>%
    mutate(
      .pred_Yes = predict(
        model,
        newdata = data,
        type = "response",
        allow.new.levels = TRUE
      ),
      .pred_No = 1 - .pred_Yes,
      .pred_class = factor(
        if_else(.pred_Yes >= 0.5, "Yes", "No"),
        levels = c("No", "Yes")
      ),
      batting_team_win = factor(
        batting_team_win,
        levels = c("No", "Yes")
      )
    ) %>%
    select(
      batting_team_win,
      .pred_No,
      .pred_Yes,
      .pred_class
    )
}

evaluate_section3_model <- function(model, train_data, test_data, model_name) {
  
  train_preds <- make_section3_preds(model, train_data)
  test_preds  <- make_section3_preds(model, test_data)
  
  train_metrics <- section3_metrics(
    train_preds,
    truth = batting_team_win,
    estimate = .pred_class,
    .pred_Yes,
    event_level = "second"
  ) %>%
    mutate(
      model = model_name,
      dataset = "Train"
    )
  
  test_metrics <- section3_metrics(
    test_preds,
    truth = batting_team_win,
    estimate = .pred_class,
    .pred_Yes,
    event_level = "second"
  ) %>%
    mutate(
      model = model_name,
      dataset = "Test"
    )
  
  train_roc <- roc_curve(
    train_preds,
    truth = batting_team_win,
    .pred_Yes,
    event_level = "second"
  ) %>%
    mutate(
      model = model_name,
      dataset = "Train"
    )
  
  test_roc <- roc_curve(
    test_preds,
    truth = batting_team_win,
    .pred_Yes,
    event_level = "second"
  ) %>%
    mutate(
      model = model_name,
      dataset = "Test"
    )
  
  list(
    metrics = bind_rows(train_metrics, test_metrics),
    roc = bind_rows(train_roc, test_roc),
    train_preds = train_preds,
    test_preds = test_preds
  )
}

section3_eval_3A <- evaluate_section3_model(
  model = model_3A_current_state_only,
  train_data = section3_train,
  test_data = section3_test,
  model_name = "Model 3A: Current State only"
)

section3_eval_3B <- evaluate_section3_model(
  model = model_3B_momentum,
  train_data = section3_train,
  test_data = section3_test,
  model_name = "Model 3B: Current State + momentum"
)

section3_performance <- bind_rows(
  section3_eval_3A$metrics
  # ,section3_eval_3B$metrics
) %>%
  select(model, dataset, .metric, .estimate) %>%
  pivot_wider(
    names_from = .metric,
    values_from = .estimate
  ) %>%
  left_join(
    section3_model_comparison %>%
      select(model, AIC, BIC, logLik),
    by = "model"
  )

section3_roc <- bind_rows(
  section3_eval_3A$roc
  # ,section3_eval_3B$roc
)

section3_performance

section3_roc %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = dataset)) +
  geom_line(linewidth = 1.1) +
  geom_abline(linetype = "dashed") +
  facet_wrap(~ model) +
  theme_minimal() +
  labs(
    title = "Section 3: Train vs Test ROC Comparison",
    x = "1 - Specificity",
    y = "Sensitivity",
    color = "Dataset"
  )
```
## Section 4

Powerplay decisions and match outcome

### Setup 

```{r section4_powerplay_setup, message=FALSE, warning=FALSE}
set.seed(42)
# Create Section 4 dataset:
# Pre-Surge powerplay = Mandatory_PP
# Surge Era powerplay = Power_Surge_1 + Power_Surge_2

df_section_4 <- df_master %>%
  mutate(
    powerplay_part = case_when(
      era == "Pre_Surge" & tactical_phase == "Mandatory_PP" ~ "Powerplay",
      era == "Surge_Era" & tactical_phase %in% c("Power_Surge_1", "Power_Surge_2") ~ "Powerplay",
      TRUE ~ "Non_Powerplay"
    )
  ) %>%
  filter(powerplay_part == "Powerplay") %>%
  mutate(
    toss_loser = case_when(
      toss_winner == team1 ~ team2,
      toss_winner == team2 ~ team1,
      TRUE ~ NA_character_
    ),
    batting_team = case_when(
      toss_decision == "bat" & innings == 1 ~ toss_winner,
      toss_decision == "bat" & innings == 2 ~ toss_loser,
      toss_decision == "field" & innings == 1 ~ toss_loser,
      toss_decision == "field" & innings == 2 ~ toss_winner,
      TRUE ~ NA_character_
    ),
    bowling_team = case_when(
      batting_team == team1 ~ team2,
      batting_team == team2 ~ team1,
      TRUE ~ NA_character_
    )
  ) %>%
  group_by(match_id, innings) %>%
  summarise(
    date = first(date),
    era = first(era),
    season = first(season),
    venue = first(venue),
    team1 = first(team1),
    team2 = first(team2),
    winner = first(winner),
    toss_winner = first(toss_winner),
    toss_decision = first(toss_decision),
    batting_team = first(batting_team),
    bowling_team = first(bowling_team),

    pp_overs = n(),
    pp_runs = sum(over_total_runs, na.rm = TRUE),
    pp_run_rate = pp_runs / pp_overs,
    pp_wickets_lost = sum(over_wickets, na.rm = TRUE),
    pp_boundaries = sum(boundaries, na.rm = TRUE),
    pp_dot_balls = sum(dot_balls, na.rm = TRUE),

    pp1_runs = sum(
      if_else(tactical_phase %in% c("Mandatory_PP", "Power_Surge_1"), over_total_runs, 0),
      na.rm = TRUE
    ),
    pp1_wickets = sum(
      if_else(tactical_phase %in% c("Mandatory_PP", "Power_Surge_1"), over_wickets, 0),
      na.rm = TRUE
    ),
    pp1_boundaries = sum(
      if_else(tactical_phase %in% c("Mandatory_PP", "Power_Surge_1"), boundaries, 0),
      na.rm = TRUE
    ),

    pp2_runs = sum(
      if_else(tactical_phase == "Power_Surge_2", over_total_runs, 0),
      na.rm = TRUE
    ),
    pp2_wickets = sum(
      if_else(tactical_phase == "Power_Surge_2", over_wickets, 0),
      na.rm = TRUE
    ),
    pp2_boundaries = sum(
      if_else(tactical_phase == "Power_Surge_2", boundaries, 0),
      na.rm = TRUE
    ),

    .groups = "drop"
  ) %>%
  mutate(
    batting_team_win = factor(
      if_else(batting_team == winner, "Yes", "No"),
      levels = c("No", "Yes")
    ),
    batting_team_win_num = if_else(batting_team_win == "Yes", 1, 0),

    era = factor(era, levels = c("Pre_Surge", "Surge_Era")),
    toss_decision = factor(toss_decision, levels = c("bat", "field")),
    batting_team = factor(batting_team),
    bowling_team = factor(bowling_team),
    season = factor(season),
    venue = factor(venue)
  ) %>%
  drop_na(
    era,
    batting_team,
    bowling_team,
    batting_team_win_num,
    pp_run_rate,
    pp_wickets_lost,
    pp_boundaries,
    pp_dot_balls
  )

# Split datasets for descriptive checking
df_section_4_pre <- df_section_4 %>%
  filter(era == "Pre_Surge")

df_section_4_surge <- df_section_4 %>%
  filter(era == "Surge_Era")

# Basic counts
df_section_4 %>%
  count(era, batting_team_win)

# Check number of powerplay overs by era
df_section_4 %>%
  count(era, pp_overs)

# Summary by era
df_section_4 %>%
  group_by(era) %>%
  summarise(
    rows = n(),
    matches = n_distinct(match_id),
    avg_pp_overs = mean(pp_overs, na.rm = TRUE),
    avg_pp_runs = mean(pp_runs, na.rm = TRUE),
    avg_pp_run_rate = mean(pp_run_rate, na.rm = TRUE),
    avg_pp_wickets = mean(pp_wickets_lost, na.rm = TRUE),
    avg_pp_boundaries = mean(pp_boundaries, na.rm = TRUE),
    avg_pp_dot_balls = mean(pp_dot_balls, na.rm = TRUE),
    batting_team_win_rate = mean(batting_team_win == "Yes", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()

# Winners vs losers by era
df_section_4 %>%
  group_by(era, batting_team_win) %>%
  summarise(
    rows = n(),
    avg_pp_run_rate = mean(pp_run_rate, na.rm = TRUE),
    avg_pp_wickets = mean(pp_wickets_lost, na.rm = TRUE),
    avg_pp_boundaries = mean(pp_boundaries, na.rm = TRUE),
    avg_pp_dot_balls = mean(pp_dot_balls, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

### Split

```{r section4_powerplay_data_split, message=FALSE, warning=FALSE}
set.seed(42)

section4_match_ids <- unique(df_section_4$match_id)

section4_train_ids <- sample(
  section4_match_ids,
  size = floor(0.8 * length(section4_match_ids))
)

section4_test_ids <- setdiff(section4_match_ids, section4_train_ids)

section4_train_raw <- df_section_4 %>%
  filter(match_id %in% section4_train_ids)

section4_test_raw <- df_section_4 %>%
  filter(match_id %in% section4_test_ids)

section4_numeric_vars <- c(
  "pp_run_rate",
  "pp_wickets_lost",
  "pp_boundaries",
  "pp_dot_balls"
)

section4_scaling_values <- section4_train_raw %>%
  summarise(
    across(
      all_of(section4_numeric_vars),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        sd = ~ sd(.x, na.rm = TRUE)
      )
    )
  )

scale_section4_data <- function(data, scaling_values) {
  data %>%
    mutate(
      pp_run_rate_z =
        (pp_run_rate - scaling_values$pp_run_rate_mean) /
        scaling_values$pp_run_rate_sd,
      
      pp_wickets_lost_z =
        (pp_wickets_lost - scaling_values$pp_wickets_lost_mean) /
        scaling_values$pp_wickets_lost_sd,
      
      pp_boundaries_z =
        (pp_boundaries - scaling_values$pp_boundaries_mean) /
        scaling_values$pp_boundaries_sd,
      
      pp_dot_balls_z =
        (pp_dot_balls - scaling_values$pp_dot_balls_mean) /
        scaling_values$pp_dot_balls_sd
    )
}

section4_train <- scale_section4_data(
  section4_train_raw,
  section4_scaling_values
)

section4_test <- scale_section4_data(
  section4_test_raw,
  section4_scaling_values
)
```

### Model
```{r section4_powerplay_model, message=FALSE, warning=FALSE}
model_4_powerplay_era <- lme4::glmer(
  batting_team_win_num ~
    era * (
      pp_run_rate_z +
        pp_wickets_lost_z +
        pp_boundaries_z +
        pp_dot_balls_z
    ) +
    toss_decision +
    (1 | batting_team) +
    (1 | bowling_team),
  data = section4_train,
  family = binomial(link = "logit")
)

summary(model_4_powerplay_era)

section4_powerplay_or <- broom.mixed::tidy(
  model_4_powerplay_era,
  effects = "fixed",
  exponentiate = TRUE,
  conf.int = TRUE
)

section4_powerplay_random <- broom.mixed::tidy(
  model_4_powerplay_era,
  effects = "ran_pars"
)

section4_model_summary <- tibble(
  model = "Model 4: Powerplay era interaction",
  AIC = AIC(model_4_powerplay_era),
  BIC = BIC(model_4_powerplay_era),
  logLik = as.numeric(logLik(model_4_powerplay_era)),
  singular = isSingular(model_4_powerplay_era)
)

fixed_4 <- fixef(model_4_powerplay_era)

section4_era_effects <- tibble(
  predictor = c(
    "pp_run_rate_z",
    "pp_wickets_lost_z",
    "pp_boundaries_z",
    "pp_dot_balls_z"
  ),
  pre_surge_effect = c(
    fixed_4["pp_run_rate_z"],
    fixed_4["pp_wickets_lost_z"],
    fixed_4["pp_boundaries_z"],
    fixed_4["pp_dot_balls_z"]
  ),
  surge_era_effect = c(
    fixed_4["pp_run_rate_z"] +
      fixed_4["eraSurge_Era:pp_run_rate_z"],
    
    fixed_4["pp_wickets_lost_z"] +
      fixed_4["eraSurge_Era:pp_wickets_lost_z"],
    
    fixed_4["pp_boundaries_z"] +
      fixed_4["eraSurge_Era:pp_boundaries_z"],
    
    fixed_4["pp_dot_balls_z"] +
      fixed_4["eraSurge_Era:pp_dot_balls_z"]
  )
) %>%
  mutate(
    pre_surge_odds_ratio = exp(pre_surge_effect),
    surge_era_odds_ratio = exp(surge_era_effect)
  )

section4_powerplay_or
section4_powerplay_random
section4_model_summary
section4_era_effects
```

### Performance 

```{r section4_powerplay_performance, message=FALSE, warning=FALSE}
section4_metrics <- metric_set(
  accuracy,
  roc_auc,
  brier_class
)

make_section4_preds <- function(model, data) {
  data %>%
    mutate(
      .pred_Yes = predict(
        model,
        newdata = data,
        type = "response",
        allow.new.levels = TRUE
      ),
      .pred_No = 1 - .pred_Yes,
      .pred_class = factor(
        if_else(.pred_Yes >= 0.5, "Yes", "No"),
        levels = c("No", "Yes")
      ),
      batting_team_win = factor(
        batting_team_win,
        levels = c("No", "Yes")
      )
    ) %>%
    select(
      batting_team_win,
      .pred_No,
      .pred_Yes,
      .pred_class
    )
}

evaluate_section4_model <- function(model, train_data, test_data, model_name) {
  
  train_preds <- make_section4_preds(model, train_data)
  test_preds <- make_section4_preds(model, test_data)
  
  train_metrics <- section4_metrics(
    train_preds,
    truth = batting_team_win,
    estimate = .pred_class,
    .pred_Yes,
    event_level = "second"
  ) %>%
    mutate(
      model = model_name,
      dataset = "Train"
    )
  
  test_metrics <- section4_metrics(
    test_preds,
    truth = batting_team_win,
    estimate = .pred_class,
    .pred_Yes,
    event_level = "second"
  ) %>%
    mutate(
      model = model_name,
      dataset = "Test"
    )
  
  train_roc <- roc_curve(
    train_preds,
    truth = batting_team_win,
    .pred_Yes,
    event_level = "second"
  ) %>%
    mutate(
      model = model_name,
      dataset = "Train"
    )
  
  test_roc <- roc_curve(
    test_preds,
    truth = batting_team_win,
    .pred_Yes,
    event_level = "second"
  ) %>%
    mutate(
      model = model_name,
      dataset = "Test"
    )
  
  list(
    metrics = bind_rows(train_metrics, test_metrics),
    roc = bind_rows(train_roc, test_roc),
    train_preds = train_preds,
    test_preds = test_preds
  )
}

section4_eval <- evaluate_section4_model(
  model = model_4_powerplay_era,
  train_data = section4_train,
  test_data = section4_test,
  model_name = "Model 4: Powerplay era interaction"
)

section4_performance <- section4_eval$metrics %>%
  select(model, dataset, .metric, .estimate) %>%
  pivot_wider(
    names_from = .metric,
    values_from = .estimate
  ) %>%
  left_join(
    section4_model_summary %>%
      select(model, AIC, BIC, logLik, singular),
    by = "model"
  )

section4_roc <- section4_eval$roc

section4_performance

conf_mat(
  section4_eval$test_preds,
  truth = batting_team_win,
  estimate = .pred_class
)

section4_roc %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = dataset)) +
  geom_line(linewidth = 1.1) +
  geom_abline(linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Section 4: Train vs Test ROC Comparison",
    x = "1 - Specificity",
    y = "Sensitivity",
    color = "Dataset"
  )
```